---
title: "Lesson 1-1.27.22"
author: "Timara Vereen"
date: "1/27/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
```{r}
install.packages("gridExtra")     
install.packages("scatterplot3d") 
install.packages("MASS")
```
```{r}
library(ggplot2)
library(gridExtra)
library(scatterplot3d)
library(MASS)
```


Univariate Assignment
Read in tree data

# read in directly from website: 

trees <- read.csv('https://raw.githubusercontent.com/dmcglinn/quant_methods/gh-pages/data/treedata_subset.csv')

# or download and import locally
trees <- read.csv('./data/treedata_subset.csv')

```{r}
trees <- read.csv('https://raw.githubusercontent.com/dmcglinn/quant_methods/gh-pages/data/treedata_subset.csv')
```

Examine this dataset and see how the data is structured, see function str

str(trees)

# we wish to model species cover across all sampled plots
# create site x sp matrix for two species 

```{r}
sp_cov = with(trees, tapply(cover, list(plotID, spcode), 
function(x) round(mean(x))))
sp_cov = ifelse(is.na(sp_cov), 0, sp_cov)
sp_cov = data.frame(plotID = row.names(sp_cov), sp_cov)
# create environmental matrix
cols_to_select = c('elev', 'tci', 'streamdist', 'disturb', 'beers')
env = aggregate(trees[ , cols_to_select], by = list(trees$plotID), 
                function(x) x[1])
names(env)[1] = 'plotID'
# merge species and enviornmental matrices
site_dat = merge(sp_cov, env, by='plotID')
# subset species of interest
abies = site_dat[ , c('ABIEFRA', cols_to_select)]
acer  = site_dat[ , c('ACERRUB', cols_to_select)]
names(abies)[1] = 'cover'
names(acer)[1] = 'cover'
```

```{r}
plot(abies)
```
```{r}
plot(acer)
```
```{r}
install.packages("car")
library(car)
```

The contents of the metadata file (./data/tree_metadata.txt) is provided here:


The dataset includes tree abundances from a subset of a vegetation database of Great Smoky Mountains National Park (TN, NC).

plotID: unique code for each spatial unit (note some sampled more than once)
date: when species occurrence recorded
plotsize: size of quadrat in m2
spcode: unique 7-letter code for each species
species: species name
cover: local abundance measured as estimated horizontal cover (ie, relative area of shadow if sun is directly above) classes 1-10 are: 1=trace, 2=0-1%, 3=1-2%, 4=2-5%, 5=5-10%, 6=10-25%, 7=25-50%, 8=50-75%, 9=75-95%, 10=95-100%
utme: plot UTM Easting, zone 17 (NAD27 Datum)
utmn: plot UTM Northing, zone 17 (NAD27 Datum)
elev: elevation in meters from a digital elevation model (10 m res)
tci: topographic convergence index, or site “water potential”; measured as the upslope contributing area divided by the tangent of the slope angle (Beven and Kirkby 1979)
streamdist: distance of plot from the nearest permanent stream (meters)
disturb: plot disturbance history (from a Park report); CORPLOG=corporate logging; SETTLE=concentrated settlement, VIRGIN=”high in virgin attributes”, LT-SEL=light or selective logging
beers: transformed slope aspect (‘heat load index’); 0 is SW (hottest), 2 is NE (coolest)


Above shows a map of the regional and local location of the elevational transects included in the dataset (from Fridley 2009).

1. Carry out an exploratory analysis using the tree dataset. Metadata for the tree study can be found here. Specifically, I would like you to develop and compare models for species cover for a habitat generalist Acer rubrum (Red maple) and a habitat specialist Abies fraseri (Frasier fir). Because this dataset includes both continuous and discrete explanatory variables use the function Anova in the packages car as such

```{r}
abies
```
```{r}
acer
```

```{r}
main_acer_mod = lm(cover ~ elev + tci + streamdist + disturb + beers, data = acer)
main_acer_mod
```
```{r}
Anova(main_acer_mod, type=3)
```
```{r}
main_abies_mod = lm(cover ~ elev + tci + streamdist + disturb + beers, data = acer)
main_abies_mod
```
```{r}
Anova(main_abies_mod, type=3)
```

```{r}
summary(main_acer_mod)
```

```{r}
summary(main_abies_mod)
```
# The p values in the summary are rounding up values, so if you wanted a more indepth analysis, the ANOVA provides more sig figs



This will estimate partial effect sizes, variance explained, and p-values for each explanatory variable included in the model.

Compare the p-values you observe using the function Anova to those generated using summary.

For each species address the following additional questions:

how well does the exploratory model appear to explain cover?

#acer: based on the anova given that we observe cover, elevation, tci, and beers are have an affect on the tree cover relative to this species
#abies: based on the anova given that we observe cover, elevation, streamdist, and disturbance have an affect on the tree over relative to this species

which explanatory variables are the most important?
#Acer: elv, tci, and beers
#Abies: elv, streamdist, and disturbance
do model diagnostics indicate any problems with violations of OLS assumptions?
#Yes...the data is discrete but the model requires continuous data and the data isn't normally distributed as well


```{r}
par(mfrow=c(3,3))
plot(main_acer_mod)
```
```{r}
par(mfrow=c(3,3))
plot(main_abies_mod)
```

```{r checking model}
par(mfrow = c(3, 3))
plot(main_abies_mod)
plot(cover ~ elev, data = abies)
plot(cover ~ tci, data = abies)
plot(cover ~ streamdist, data = abies)
plot(cover ~ beers, data = abies)
```

are you able to explain variance in one species better than another, why might this be the case?

#it's better to explain variance with generalized model than specialist model because with a glm, it's more bi-modal distributed and can see this distribution of data while the specialist model 
Prior to addressing the above questions you will want to restructure and subset the data using the following R code:

# we wish to model species cover across all sampled plots
# create site x sp matrix for two species 
sp_cov = with(trees, tapply(cover, list(plotID, spcode), 
                           function(x) round(mean(x))))
sp_cov = ifelse(is.na(sp_cov), 0, sp_cov)
sp_cov = data.frame(plotID = row.names(sp_cov), sp_cov)
# create environmental matrix
cols_to_select = c('elev', 'tci', 'streamdist', 'disturb', 'beers')
env = aggregate(trees[ , cols_to_select], by = list(trees$plotID), 
                function(x) x[1])
names(env)[1] = 'plotID'
# merge species and enviornmental matrices
site_dat = merge(sp_cov, env, by='plotID')
# subset species of interest
abies = site_dat[ , c('ABIEFRA', cols_to_select)]
acer  = site_dat[ , c('ACERRUB', cols_to_select)]
names(abies)[1] = 'cover'
names(acer)[1] = 'cover'

2\. You may have noticed that the variable cover is defined as 
positive integers between 1 and 10. and is therefore better treated
as a discrete rather than continuous variable. 
Re-examine your solutions to the question above but from the
perspective of a General Linear Model (GLM) with a Poisson error term
(rather than a Gaussian one as in OLS). 
The Poisson distribution generates integers 0 to positive infinity so this may provide a good first approximation. 
Your new model calls will look as follows:

acer_poi = glm(cover ~ tci + elev + ... , data = my_data, 
           family='poisson')
For assessing the degree of variation explained you can use a pseudo-R-squared statistic (note this is just one of many possible)

```{r}
pseudo_r2 = function(glm_mod) {
                1 -  glm_mod$deviance / glm_mod$null.deviance
}
```

```{r}
pseudo_r2 = function(glm_mod) {
  1- glm_mod$deviance / glm_mod$null.deviance
}
pseudo_r2(acer_poi)
pseudo_r2(abies_poi)
```
# the R-squared values for the acer_poi and abies_poi are both positive, yet are different in size displaying that the acer_poi had a weaker correlation in comparison to the abies_poi which displayed a higher correlation

 Compare your qualatitive assessment of which variables were most important in each model. Does it appear that changing the error distribution changed the results much? In what ways?
 #in relation to the p-values The poisson test is more restrictive and sensitive when compared to the general linear model
```{r}
acer_poi = glm(cover ~ elev + tci + streamdist + disturb + beers, data = acer, 
           family='poisson')
acer_poi
#main_aCER_mod = lm(cover ~ elev + tci + streamdist + disturb + beers, data = acer)
```
```{r}
abies_poi = glm(cover ~ elev + tci + streamdist + disturb + beers, data = abies, 
           family='poisson')
abies_poi
#main_aCER_mod = lm(cover ~ elev + tci + streamdist + disturb + beers, data = acer)
```

3. Provide a plain English summary (i.e., no statistics) of what you have found and what conclusions we can take away from your analysis?
#From our analysis, we can tell when we look at error distributions, we may receive different p-values relative to our levels of the study. Given the poisson test, our p-values were more sensitive in comparison to the glm creating more significance for the data.


```{r}
trees <- read.csv('https://raw.githubusercontent.com/dmcglinn/quant_methods/gh-pages/data/treedata_subset.csv')
```




